---
title: "faissR: nearest-neighbor search, prediction, and k-means"
author: "Stefano Cacciatore"
package: faissR
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
header-includes:
  - \usepackage{array}
vignette: >
  %\VignetteIndexEntry{faissR: nearest-neighbor search, prediction, and k-means}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 6,
    fig.height = 4
)
```

# Overview

`faissR` provides native nearest-neighbour search, k-nearest-neighbour
prediction, and k-means utilities for R
workflows that use large row-wise matrices. The package is designed for
high-throughput biological data analysis, including single-cell, flow
cytometry, imaging, and mass spectrometry applications.

The package requires the FAISS C++ library for CPU indexes. CUDA, FAISS GPU
indexes and NVIDIA cuVS are optional. CPU-only
installations do not need NVIDIA libraries, and explicit CUDA requests fail
clearly when CUDA support is not available.

The main public functions are:

| Function | Purpose |
|---|---|
| `nn()` | nearest-neighbour search on host matrices |
| `nn_gpu()` | GPU-resident exact-family KNN result for downstream CUDA code |
| `candidate_knn()` | refine a supplied candidate neighbour set |
| `knn()` and `predict()` | supervised kNN fitting and prediction |
| `fast_kmeans()` | CPU or optional CUDA k-means |
| `backend_info()` and `nn_capabilities()` | inspect compiled backends |
| `print()` and `summary()` | inspect result and fitted-model metadata |

# Installation contract

FAISS is mandatory. A source installation stops when its headers or library
cannot be found; it does not silently substitute an unrelated nearest-neighbor
implementation. On macOS, a typical system installation is
`brew install faiss libomp`. On Linux, install the distribution's FAISS
development package or set `FAISS_HOME` to a prefix containing
`include/faiss` and `lib/libfaiss`. CUDA and cuVS are optional unless a GPU
build is explicitly requested.

Automatic Homebrew installation is never enabled by default. GitHub users who
explicitly want that convenience can set
`FAISSR_AUTO_INSTALL_FAISS=1` before installation. Unsupported WebAssembly and
package-staging platforms can compile diagnostic stubs for portability checks;
those artifacts report that native search is unavailable and are not
functional FAISS builds.

# Example data

The examples use a small synthetic matrix with three groups. Rows are
observations and columns are features, matching the input convention used by
`faissR`.

```{r data}
library("faissR")

set.seed(100)
n_per_group <- 30L
p <- 8L

centers <- rbind(
    rep(-2, p),
    rep(0, p),
    rep(2, p)
)

x <- do.call(rbind, lapply(seq_len(nrow(centers)), function(i) {
    sweep(
        matrix(rnorm(n_per_group * p, sd = 0.7), ncol = p),
        2,
        centers[i, ],
        "+"
    )
}))

groups <- factor(rep(LETTERS[1:3], each = n_per_group))
dim(x)
table(groups)
```

# Backends

`backend_info()` reports which compiled routes are available in the current R
session. On a CPU-only machine, CUDA-related entries are expected to be
unavailable.

```{r backends}
available_backends <- backend_info()
available_backends <- available_backends[
    c("backend", "available", "knn_available")
]
available_backends
```

`nn_capabilities()` gives method, backend, and metric support. The table can be
large, so the example below shows only CPU Euclidean rows.

```{r capabilities}
caps <- nn_capabilities()
subset(
    caps,
    backend == "cpu" & metric == "euclidean",
    select = c("method", "backend", "metric", "supported", "exact")
)
```

# Nearest-neighbour search

The main entry point is `nn()`. Use `exclude_self = TRUE` for self-KNN when
each row is searched against the same matrix and should not return itself.

```{r nn-exact}
nearest <- nn(
    x,
    k = 5,
    backend = "cpu",
    method = "exact",
    metric = "euclidean",
    exclude_self = TRUE,
    n_threads = 2
)

nearest
summary(nearest)
dim(nearest$indices)
head(nearest$indices)
head(round(nearest$distances, 3))
```

The returned object stores both the public request and the implementation that
actually ran.

```{r nn-metadata}
c(
    requested_backend = attr(nearest, "requested_backend"),
    requested_method = attr(nearest, "requested_method"),
    backend_used = attr(nearest, "backend_used"),
    metric = attr(nearest, "metric"),
    distance_type = attr(nearest, "distance_type"),
    exact = attr(nearest, "exact"),
    exclude_self = attr(nearest, "exclude_self")
)
```

`exact`, `flat`, and `bruteforce` all request exhaustive mathematics, but they
express different provider intentions. `exact` lets faissR choose an exhaustive
provider, `flat` explicitly requests FAISS Flat, and CUDA `bruteforce` prefers
direct cuVS brute force. They should have the same neighbor quality; timing
differences measure provider and wrapper overhead.

## Automatic method selection

The default `method = "auto"` uses compiled shape-, metric-, `k`-, backend-,
and target-recall-aware rules. These rules are deterministic; they do not run a
slow tuning pass inside ordinary user calls.
For approximate methods, `target_recall` selects a calibrated operating policy;
it is not a guarantee for a new dataset. Inspect
`tuning_benchmark_target_met` in the returned metadata and validate recall
against an exact reference when the application requires a quality bound.

```{r nn-auto}
auto_nearest <- nn(
    x,
    k = 8,
    backend = "cpu",
    method = "auto",
    metric = "euclidean",
    exclude_self = TRUE,
    n_threads = 2
)

c(
    backend_used = attr(auto_nearest, "backend_used"),
    requested_method = attr(auto_nearest, "requested_method")
)
```

For large two- or three-dimensional self-KNN, `method = "auto"` can select the
native grid route. CPU and CUDA have separate implementations. The explicit
CPU call below demonstrates the API on a small two-dimensional matrix.

```{r grid}
x2 <- matrix(runif(2000), ncol = 2)
grid_nearest <- nn(
    x2,
    k = 10,
    backend = "cpu",
    method = "grid",
    metric = "euclidean",
    exclude_self = TRUE,
    n_threads = 2
)

c(
    backend_used = attr(grid_nearest, "backend_used"),
    k = attr(grid_nearest, "k")
)
```

CUDA uses its native grid implementation. Explicit CUDA calls fail when the
requested route is unavailable and never report a CPU fallback as GPU work.

```{r cuda-grid, eval=FALSE}
cuda_nearest <- nn(
    x2,
    k = 10,
    exclude_self = TRUE,
    backend = "cuda",
    method = "grid",
    metric = "euclidean"
)
```

The same failure rule applies before launch. For example, CAGRA is CUDA-only:

```{r unsupported-combination}
try(
    nn(
        x,
        k = 5,
        backend = "cpu",
        method = "cagra",
        exclude_self = TRUE
    )
)
```

# Metrics

The public metric set is intentionally small:

- `euclidean`;
- `cosine`;
- `correlation`;
- `inner_product`.

Cosine normalizes each row before searching. Correlation first centers each row
and then applies row normalization. Inner product ranks neighbours by larger
raw dot product, while returned distances keep faissR's smaller-is-better
convention.

```{r metrics}
metric_results <- lapply(
    c("euclidean", "cosine", "correlation", "inner_product"),
    function(metric) {
        out <- nn(
            x,
            k = 5,
            backend = "cpu",
            method = "exact",
            metric = metric,
            exclude_self = TRUE,
            n_threads = 2
        )
        data.frame(
            metric = metric,
            backend_used = attr(out, "backend_used"),
            first_distance = round(out$distances[1, 1], 4)
        )
    }
)

do.call(rbind, metric_results)
```

# Float32 input and output

FAISS and cuVS consume float32 data internally. Ordinary R numeric matrices are
converted once for those compiled routes. If the optional `float` package is
installed, `faissR` can also accept `float::fl()` input and return float32
distance matrices with `output = "float"`.

```{r float32, eval=requireNamespace("float", quietly = TRUE)}
xf <- float::fl(x)

float_nearest <- nn(
    xf,
    k = 5,
    backend = "cpu",
    method = "flat",
    metric = "euclidean",
    exclude_self = TRUE,
    output = "float",
    n_threads = 2
)

c(
    input_type = attr(float_nearest, "input_type"),
    distance_type = attr(float_nearest, "distance_type")
)
class(float_nearest$distances)
```

# Reusing fitted indexes with kNN models

`knn()` can fit a model for repeated prediction, or it can fit and predict in
one call when `Xtest` is supplied. Fitted FAISS Flat, HNSW, IVF, and IVFPQ
indexes are reused by `predict()` where supported.

```{r knn-fit}
fit <- knn(
    Xtrain = x,
    Ytrain = groups,
    k = 7,
    backend = "cpu",
    method = "exact",
    metric = "euclidean",
    n_threads = 2
)

pred <- predict(fit, x[1:6, , drop = FALSE])
prob <- predict(fit, x[1:6, , drop = FALSE], type = "prob")

summary(fit)
as.vector(pred)
prob_display <- matrix(
    as.numeric(prob),
    nrow = nrow(prob),
    dimnames = dimnames(prob)
)
round(prob_display, 3)
```

Immediate prediction uses the same interface:

```{r knn-immediate}
immediate <- knn(
    Xtrain = x,
    Ytrain = groups,
    Xtest = x[1:6, , drop = FALSE],
    k = 7,
    backend = "cpu",
    method = "exact",
    metric = "euclidean",
    n_threads = 2
)

as.vector(immediate)
```

# Candidate refinement

`candidate_knn()` ranks only the supplied candidate rows. This is useful when
another algorithm constructs a graph or a broad candidate pool and faissR
should apply the final metric consistently.

```{r candidate-refinement}
candidates <- nearest$indices
refined <- candidate_knn(
    data = x,
    points = x,
    candidates = candidates,
    k = 3,
    metric = "euclidean",
    exclude_self = TRUE,
    n_threads = 2
)
summary(refined)
```

# k-means

`fast_kmeans()` uses the same backend style as the nearest-neighbour functions.
The CPU route is available everywhere; CUDA is optional.

```{r kmeans}
km <- fast_kmeans(
    x,
    centers = 3,
    backend = "cpu",
    n_init = 2,
    max_iter = 20
)

summary(km)
c(
    backend = km$backend,
    backend_library = km$backend_library,
    converged = km$converged
)
table(km$cluster, groups)
```

# Optional CUDA use

CUDA-specific calls should be guarded with `cuda_available()` or inspected with
`backend_info()`. The example below is not evaluated on CPU-only builders.

```{r cuda-example, eval=FALSE}
if (cuda_available()) {
    gpu_knn <- nn_gpu(
        x,
        k = 15,
        exclude_self = TRUE,
        method = "auto",
        metric = "euclidean",
        tuning = "auto",
        target_recall = 0.99
    )

    gpu_knn
    host_copy <- gpu_knn_to_host(gpu_knn)
}
```

`nn_gpu()` is for downstream CUDA consumers. It keeps result buffers resident on
the GPU and copies them back to R only when `gpu_knn_to_host()` is called
explicitly. Its current public method set is intentionally limited to
`"auto"`, `"exact"`, `"flat"`, and `"bruteforce"`. Euclidean inputs with more
than three columns and raw inner product use the FAISS GPU direct brute-force
primitive when available. Two- and three-dimensional Euclidean inputs use a
native direct-difference CUDA kernel to avoid float32 cancellation for nearly
coincident vectors. Cosine and correlation use the native CUDA exact route
after their documented row transformations. If ordinary
`nn(method = "auto", backend = "cuda")` would
prefer an approximate method, the GPU-resident object records that preference
and the exact-family residency constraint instead of silently claiming that
the approximate route executed.

## Compiled inter-package API

CUDA-enabled downstream R packages can consume the same device-resident result
without calling an R wrapper or copying through host matrices. They should add
`faissR` to `LinkingTo`, include the installed `<faissR_api.h>` header, check
that `faissR_c_api_version()` returns the ABI version they support, and obtain
the typed function pointer with `faissR_get_nn_cuda_tuned_gpu()`. The callable
returns the same owning `faissR_gpu_knn` external-pointer object as `nn_gpu()`.
The R object must remain protected while downstream C++ or CUDA code uses its
non-owning device pointers; only `gpu_knn_to_host()` performs an explicit host
copy. The header also provides typed accessors for the host-returned float32
nearest-neighbour callables.

# Reproducibility and third-party software

Every nearest-neighbour result records the requested and resolved route,
metric, exactness, self-exclusion policy, input/output precision, and available
provider-specific tuning metadata. These fields should be archived with
benchmark results rather than inferred later from a method label.

The publication benchmark launchers additionally pin the package version and
40-character source commit. Raw timing rows and exact-reference objects record
both the required package commit and the commit embedded in the container; the
freeze audit rejects missing or unequal identities before aggregation.
The repository's phase-aware HPC submitter preserves one method/backend/metric
per Slurm launcher and persists its ledger after every submission, including
partial phases. The compact JSS replication script remains a separate CPU path
that can be run on an ordinary computer.

`faissR` is MIT licensed. FAISS, CUDA, and cuVS remain external system
libraries under their own licenses. The package includes a small adaptation of
the public DLPack C ABI definitions under Apache License 2.0; its installed
third-party notice and full license are available through:

```{r third-party-notices}
stopifnot(nzchar(system.file("THIRD-PARTY-NOTICES", package = "faissR")))
stopifnot(nzchar(system.file("licenses", "DLPACK-LICENSE.txt", package = "faissR")))
cat("Third-party notices installed: TRUE")
```

# Session information

```{r session-info}
sessionInfo()
```
