---
title: "Installing faissR and auditing native capabilities"
author: "Stefano Cacciatore"
package: faissR
output:
    BiocStyle::html_document:
        toc: true
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Installing faissR and auditing native capabilities}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Build contract

A functional `faissR` installation requires R development headers, a C++20
compiler, a Fortran compiler, and ABI-compatible FAISS headers and libraries.
CUDA, FAISS GPU, and RAPIDS cuVS are optional. They become mandatory only when
their strict build switches are set. The package does not vendor these native
libraries and never installs a system package manager dependency silently.

Use `FAISSR_REQUIRE_FAISS=1` for a production or CI build. This makes a missing
or unusable FAISS installation fatal. A *diagnostic-only* build can be emitted
on explicitly recognized automated targets that do not provide FAISS. Such an
artifact can load and explain the missing capability, but it cannot perform a
nearest-neighbour search and is not a functional platform result.

# Tested combinations

| Purpose | Platform and R | Native providers | Compiler information |
|---|---|---|---|
| Functional CPU install and package check | macOS arm64; R 4.6.0 | FAISS 1.14.3; libomp 22.1.x | Homebrew clang 22.1.1; GNU Fortran 12.2.0 |
| Publication CPU/CUDA campaign | Debian 13; R 4.5.3; NVIDIA L40S, driver 595.58.03 | FAISS 1.14.3; cuVS 26.06; CUDA 13.2 | Compiler executable/version was not retained in the frozen image |
| Repository CPU CI definition | Ubuntu; current Bioconductor devel R | FAISS 1.14.3 built from source | Hosted GCC/GFortran toolchain recorded by the job |
| Repository CPU CI definition | macOS; current Bioconductor devel R | Homebrew FAISS and libomp | Hosted clang/GFortran toolchain recorded by the job |

The CI rows describe tracked workflows. They become release evidence only
after the corresponding hosted jobs complete successfully.

# Linux CPU-only installation

Ubuntu releases do not all provide `libfaiss-dev`. The following reproducible
path builds FAISS 1.14.3 without CUDA, Python bindings, or tests and installs it
under `/opt/faiss`.

```{bash, eval=FALSE}
sudo apt-get update
sudo apt-get install -y build-essential cmake ninja-build gfortran \
    git libopenblas-dev liblapack-dev

git clone --depth 1 --branch v1.14.3 \
    https://github.com/facebookresearch/faiss.git faiss-1.14.3
cmake -S faiss-1.14.3 -B faiss-1.14.3/build -G Ninja \
    -DFAISS_ENABLE_GPU=OFF -DFAISS_ENABLE_PYTHON=OFF \
    -DBUILD_TESTING=OFF -DBUILD_SHARED_LIBS=ON \
    -DCMAKE_BUILD_TYPE=Release -DCMAKE_INSTALL_PREFIX=/opt/faiss
cmake --build faiss-1.14.3/build --parallel
sudo cmake --install faiss-1.14.3/build

FAISS_HOME=/opt/faiss \
FAISSR_REQUIRE_FAISS=1 \
FAISSR_USE_CUDA=0 FAISSR_USE_CUVS=0 FAISSR_USE_CUGRAPH=0 \
R CMD INSTALL faissR_0.99.29.tar.gz
```

When the distribution supplies compatible development files, installing
`libfaiss-dev` and omitting `FAISS_HOME` is sufficient.

# macOS CPU-only installation

The tested Apple Silicon path is:

```{bash, eval=FALSE}
brew install faiss libomp gcc

FAISS_HOME="$(brew --prefix faiss)" \
LIBOMP_HOME="$(brew --prefix libomp)" \
FAISSR_REQUIRE_FAISS=1 \
R CMD INSTALL faissR_0.99.29.tar.gz
```

An already active conda/mamba environment is also accepted when it contains
`faiss-cpu` and `libomp`; set both `FAISS_HOME` and `LIBOMP_HOME` to
`$CONDA_PREFIX`. `configure` does not create or modify the environment.

# Linux CUDA and cuVS installation

The CUDA toolkit, driver, FAISS GPU build, cuVS, R, and C++ runtime must come
from one compatible stack. After those providers have been installed in known
prefixes, use strict switches so that a CPU-only build cannot be mistaken for
the requested GPU build:

```{bash, eval=FALSE}
CUDA_HOME=/usr/local/cuda \
FAISS_HOME=/opt/faiss-gpu \
CUVS_HOME=/opt/rapids \
FAISSR_REQUIRE_FAISS=1 \
FAISSR_REQUIRE_CUDA=1 \
FAISSR_REQUIRE_CUVS=1 \
R CMD INSTALL faissR_0.99.29.tar.gz
```

Use `FAISSR_REQUIRE_CUGRAPH=1` as well only when graph-clustering support is
required. Native RAPIDS builds are Linux-oriented; WSL2 is the documented
Windows route for CUDA/cuVS.

# Audit the installed capability

Loading is not enough to establish a functional native installation. Inspect
the provider table and run a small CPU search:

```{r capability-audit}
library(faissR)

backend_info()
head(nn_capabilities(runtime = TRUE))

stopifnot(faiss_available())
x <- matrix(seq_len(240), ncol = 6)
ans <- nn(x, k = 3, method = "exact", backend = "cpu")
stopifnot(identical(dim(ans$indices), c(nrow(x), 3L)))

c(
    faiss_gpu = faiss_gpu_available(),
    cuda = cuda_available(),
    cuvs = cuvs_available()
)
```

Do not infer GPU support from `nvidia-smi` alone. The package must have been
compiled against the provider and the runtime device must also be available.

# Common failures

| Message or symptom | Interpretation | Resolution |
|---|---|---|
| FAISS headers/library not found | No compatible FAISS prefix was discovered | Set `FAISS_HOME`; verify `include/faiss/Index.h` and `lib/libfaiss.*`. |
| `omp.h` not found on macOS | FAISS headers require an OpenMP runtime not visible to clang | Install `libomp` and set `LIBOMP_HOME`. |
| Package loads but `faiss_available()` is false | This is a diagnostic-only artifact | Install from source with FAISS and `FAISSR_REQUIRE_FAISS=1`. |
| `libfaiss` or `GLIBCXX_*` load failure | Runtime linker/compiler ABI mismatch | Use one compiler/runtime stack and expose its library directory through the platform linker path. |
| CUDA requested but `nvcc` is missing | CUDA toolkit was not discovered | Set `CUDA_HOME`/`NVCC` and check `nvcc --version`. |
| cuVS route unavailable | cuVS was absent at compile time or cannot load at runtime | Set `CUVS_HOME`, use `FAISSR_REQUIRE_CUVS=1`, and repair the runtime library path. |

# Checks and licenses

The source tarball should be checked in the same strict environment used for
installation:

```{bash, eval=FALSE}
R CMD build .
FAISS_HOME=/opt/faiss FAISSR_REQUIRE_FAISS=1 \
    R CMD check --as-cran faissR_0.99.29.tar.gz
Rscript -e 'BiocCheck::BiocCheck("faissR_0.99.29.tar.gz", `new-package`=TRUE)'
```

`faissR` is MIT-licensed. FAISS is MIT-licensed, cuVS is Apache-2.0-licensed,
and CUDA is governed by the NVIDIA CUDA SDK EULA. These external components
are supplied by the system and are not redistributed in the package. The
installed `THIRD_PARTY_LICENSES.md` records links and the distribution
boundary.

# Session information

```{r session-info}
sessionInfo()
```
